\large \textbf{Leveraging Historical Ties between Cognitive Science and Computer Science to Guide Programming Education}

### 2016-05-03 paper.tex
no need for relative path prefix
final proofread and edits

% sorting=none will order citations in order of appearance
% use sorting=nyt to sort by name, year, title
\usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
\addbibresource{./computing.bib}
\appto{\bibfont}{\footnotesize} % for 10pt bibliography

@@

\item Even if goals were clearly stated it is likely that many
  different approaches could be used successfully if chosen
  intentionally to align with the goals and measurements of success
  (assessment). In many of the studies, it is not immediately clear if
  the measures of success used are appropriate for the stated
  objectives. That is to say, the assessments used might not always be
  measuring all aspects of the desired learning objectives.
\end{enumerate}

@@

It should be strongly noted that we are not advocating for OOP as a
``best'' teaching paradigm. Rather, we want to point out that from
what has been reported, teaching OOP effectively can be
challenging. What many implementations seem to miss is that OOP
defines an entirely different paradigm for problem solving than is
used in procedural programming\autocite{kolling_problem_1999}. As
should not be surprising there are some problems for which OOP
provides a good set of tools to solve in an efficient, elegant manner,
namely those with solutions that naturally map to real-world
objects\autocite{raymond_art_2003}. Different problems may be better
solved using a procedural, functional, or data-flow
paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
our goal should be to help students analyze problem structures and
make informed decisions regarding what paradigm would be best for a
particular problem. What is not clear from the literature is whether
or not this consideration is made when designing problems for a
particular course. Indeed, in some cases there is evidence that OOP
concepts are added to a course ``on top of'' older content, without
necessarily changing the problem
sets\autocite{kolling_problem_1999,vilner_fundamental_2007}.

\subsection*{Cognitive load theory}
Cognitive load theory (CLT) is based on the premise that the capacity
of our working memory, where all conscious cognitive processing takes
place, is finite and relatively small: we are able to manage 2-3
interacting elements at a time. Examples of ``interacting elements''
in the current context would be recalling and using the syntax and
grammar of a particular programming language, reasoning about the
behavior of a single object in an OOP environment, or reasoning about
a solution to a particular high-level problem. CLT posits that in
light of this finite capacity of working memory, for learning to
occur, all aspects of a task must fit into this finite space. While
some contributors to cognitive load are necessary for the task at
hand, others are not\autocite{paas_cognitive_2003}. Cognitive load can
be divided into three different categories: \emph{intrinsic} cognitive
load is due to the complexity of the task itself and generally can not
be reduced without simplifying the task. \emph{Extraneous} cognitive
load is associated with work that is due to the instructional method,
but not part of the task at hand. For instance, if the educational
goal is to complete a programming task, but learners must search and
interpret reference material in order to do so, the search and
interpretation of this material would be considered extraneous
cognitive load. Finally, \emph{germane} cognitive load is also a
result of the instructional design, but unlike extraneous load,
germane load enhances the learning process. For instance, if a
learning goal is that students are able to navigate and use reference
material, than the previous example would be considered germane
cognitive load. What this example highlights is that whether
complexity due to the instruction design is extraneous or germane
depends on the specific learning goals. The role of the instructional
designer then is to reduce extraneous load and intentionally managing
germane load.

\subsection*{Object oriented programming}
%% Reviewer: more explicit connections to cogsci
Alan Kay originally conceived of object oriented programming as
\emph{one} way to think about complex
systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
conception for OOP was the biological cell: each object is
designed to accomplish a specific set of tasks where the complexity of
these tasks is encapsulated within the object. Objects interact with
one another by sending and receiving messages, and systems of these
objects are assembled to solve complex
problems\autocite{kay_history_1996}.

@@

Ultimately, if our goal is to teach problem solving in general, we
would be doing a grave disservice to our students to leave them with
the impression that there is only one paradigm, whether it be OOP,
procedural, functional, or something not yet invented. As previously
mentioned, OOP and procedural paradigms are comprised of fundamentally
different concepts for problem solving, and so again is the functional
paradigm. Unsurprisingly, there is no paradigm that is the ``best''
fit for every problem. Leaving students with the belief that only one
paradigm exists makes it very difficult for them to learn how to
intentionally select and apply the ``best'' paradigm to solve a
particular problem. \citeauthor{kolling_problem_1999} argues that
performing a paradigm shift from procedural style to OOP is difficult,
and so concludes that if OOP is a desired outcome then it should be
taught \emph{first} since the reverse shift, from OOP to procedural,
is conceptually easier. What this recommendation neglects to account
for however is that both paradigms depend upon authoring programs in
some language, and this authoring process takes a certain amount of
cognitive capacity. To justify an objects-first approach we would have
to verify that the cognitive load associated with OOP concepts
themselves combined with the cognitive load of learning to author a
program in a particular environment, do not exceed the learner's
working memory capacity.

In a superficial way, CLT theory would suggest that OOP is a helpful
paradigm for managing cognitive omplexity: objects encapsulate
complexity by hiding the details of implementation behind a simple
interface. For example, an object that represents a list of items may
have operations to add and remove items from the list. Conceptually we
can use this object with our common understanding of what it means to
``add'' an ``item'' to a ``list''. Critically, we do \emph{not} need
to be aware of how the adding and removal of the item is actually
implemented in code, or even how the storage is structured in memory
(e.g. an array or a linked-list). We refer to this use of CLT to
justify the reduction of complexity of OOP as superficial because it
focuses on only the conceptual paradigm itself without regard to
implementing it in some programming environment. In practice, we
usually ask our students to implement OOP concepts in a particular
programming language. That task requires not only understanding OOP as
a paradigm, but also working with the programming environment,
authoring a program, understanding the syntax and grammar of the
language, and most likely debugging it as well.

\subsection*{Cognitive challenges teaching OOP}
A behavior that characterizes expert programmers is the ability to
easily shift between different abstraction layers of a problem
space\autocite{armoni_teaching_2013}. This ability may be especially
crucial to becoming fluent in OOP as the concepts are somewhat further
removed from the actual syntax than in other paradigms. For example,
in the procedural paradigm the central concept is that of the
procedure, or subroutine. Subroutines are simply collections of
imperative commands grouped together under a label. Thus, if someone
can write and understand imperative code, the conceptual leap to
understanding the procedural paradigm is relatively small. In
contrast, to understand objects in OOP one must understand the concept
of state and methods operating on the state. Both state and method are
implemented with code (generally imperative), but to understand how an
object functions one has to imagine its initial state, and then how
each method call would change this state. Cognitive load theory would
suggest that understanding the dynamic interaction of state and method
calls is more challenging than grouping lines of imperative codes into
re-usable procedures. Thus to be fluent in OOP design a programmer
\emph{must} be able to switch between thinking at the object layer and
the program layer. Teaching abstraction is best done by doing it by
example, while being explicit about the abstractions
used\autocite{armoni_teaching_2013}. However, we must keep in mind the
limited capacity of working memory: if the cognitive load associated
with writing code and thinking about state and methods fills up a
learners working memory they may have difficult learning concepts of
abstraction, even when the instructor explicitly points them out. To
reduce the cognitive load associated with writing code and working
with object state and methods, these tasks must be practiced until
somewhat automated: they can be done without conscious
thought\autocite{ambrose_chapter_2010}.

\subsection*{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. The
explanation for this can be found in CLT. If, for example, the goal is
to teach problem-solving techniques, there will be cognitive load
associated with thinking about different strategies one could
use. Learning the syntax rules of a programming language adds to the
cognitive load, but it does not aid in the process of learning
problem-solving techniques. In other words, learning the syntax of a
programming language is an extraneous cognitive load in this
context. Thus, it is in the educator's best interest to minimize
cognitive load associated with features that do not directly help with
the learning objectives of the course and select a teaching language
with easy-to-learn syntax.

@@

In theory an iconic programming environment such as
Scratch\autocite{resnick_scratch_2009} or
Alice\autocite{cooper_using_2003} should aid in the learning of higher
level concepts, such as those associated with OOP, since graphical
languages eliminates the extraneous cognitive load of dealing with
syntax and grammar of a language. In fact, there has been success in
using iconic languages to teach first year programming
\autocite{calloni_iconic_1997}, and research indicates that teaching
graphical programming first does not inhibit a student from later
transferring learned high level concepts to textual programming
languages\autocite{hundhausen_can_2009}.

\citeauthor{snook_incorporation_2005} describe efforts to use the Alice
programming environment to teach introductory programming
concepts\autocite{snook_incorporation_2005}. Interestingly a later
report assessing the efficacy of the curriculum using Alice indicated
that while pre-post testing indicated learning gains with the
environment, focus group data indicated a dissatisfaction with Alice
as an introductory language resulting in a switch to
LabVIEW\autocite{lohani_reformulating_2011}.

\subsection*{Problem solving}
Developing problem solving was widely regarded as a common objective
for introductory programming courses. Achieving this objective is more
likely if the course begins with a focus on problem solving
strategies, before any programming is
introduced\autocite{koulouri_teaching_2014}. Separating
problem-solving instruction in this way has a number of benefits
grounded in cognitive science research: first, it serves to activate
prior knowledge\autocite{ambrose_how_2010}. With problem-solving
strategies fresh in their memory, learners will more easily apply them
when they begin programming. Second, it helps manage cognitive
load. Where as in the previous example of language choice we reduced
cognitive load by choosing a language with a simpler syntax, here we
separate two tasks that we know individually demand a high level of
cognitive load for novices: learning problem-solving strategies, and
learning programming techniques.

@@

\section*{Assessment}
%% Reviewer: short
Often times assessment becomes the weak point of a learning
system. Designing good assessment is
hard\autocite{lister_blooming_2000,traynor_synthesis_2005}, and often good assessment can
require significant time on the part of the instructor. Because of
this, assessment is often the weak point, lowering the effectiveness
of the entire system.

If our course objective is to teach learners to solve problems with
programming then an ideal assessment would be to give students a
problem to solve, ask them to write a program to solve it as well as
reflect upon the patterns they used in their solution. Asking for
reflection prompts students to engage in meta-cognition, becoming
aware of their own thinking process, which helps them become more
effective
learners\autocite{ambrose_how_2010,chi_selfexplanations_1989}. Of
course evaluating this type of assessment is much more time consuming
than a multiple-choice exam that can be automatically scored, or even
than evaluating that the finished program runs correctly, which can be
done automatically as well\autocite{cheang_automated_2003}.

\section*{Considerations and limitations}
In the above hypothetical example we had the luxury of complete
freedom in designing all aspects of an introductory programming
curriculum. We recognize this is almost always never the case in
practice; however, we argue that following a framework, such as
constructive alignment, as close as possible will help. Thinking about
the framework will also encourage critical engagement with the
constraints imposed by the institution. For example, there has been
much interest in assessment tools used for introductory programming
courses\autocite{lister_blooming_2000,lister_objectives_2001,sheard_exploring_2011},
and in particular automated
assessment\autocite{cheang_automated_2003,traynor_synthesis_2005,ihantola_review_2010}. In
many cases the push towards automated assessment techniques is a
response to increasing class sizes that instructors typically have
little control over. In the event that a large class size dictates
that some form of automated assessment be used, following the
constructive alignment framework would constrain the types of course
objectives and instruction that were feasible. For instance, if the
primary course objective was to teach conceptual understanding of
computer science, then a well-designed multiple-choice assessment
could be used
effectively\autocite{lister_blooming_2000,lister_objectives_2001}.

@@

An area that has not been addressed in this paper but has strong
connections is that of attention in
cognition\autocite{simon_role_1986}. There are strong parallels
between our limitations to focus our conscious attention at one or two
tasks at a time, and the serial processing inherent in most computing
and language design. Of particular interest is the move in hardware to
increasingly parallel designs to keep pace with Moore's law. Our
cognitive constraints of attention may suggest an inherent difficulty
in thinking and reasoning about parallel processing that must be
addressed if we want to prepare students for an environment that
demands parallel-processing solutions to programming problems. These
demands may shift focus to languages such as Erlang that are designed
for high levels of concurrency as well as demand we adopt different
metaphors for computing\autocite{stein_challenging_1999}.

@@

Recognizing that introductory programming is a common component of
engineering curriculum, and that there is a national push to introduce
it to an even larger population, it is important to think carefully
about how we design programming education to impart the cognitive
skills that will generally benefit learners even after the specific
technologies they learn are no longer relevant. We strongly advocate
for the use of a framework such as constructive alignment, described in
this paper, to inform the choices for course objectives, instructional
techniques, and assessments. Using such a framework well help identify
components of a course that may be contributing unnecessarily to
cognitive load, or reasons why student outcomes might not match
expectations assumed from course objectives.

Analyzing a proposed curriculum through the lens of cognitive load
theory can aid in instruction decisions that will increase the chances
achieving learning goals. Educators must critically review the goals
of their introductory programming courses: if goals are to understand
high level concepts such as OOP, but not necessarily implement these
concepts in textual languages, it may be prudent to consider iconic
languages for these courses. If however, goals include learning OOP
concepts \emph{and} implementing these concepts in textual language,
educators must be aware of the increased cognitive load associated
with each of these tasks and provide scaffolding to help students
automate aspects of each task individually before expecting them to
combine them both in successful programs.

Taken together, using an established framework, such as constructive
alignment, to design an analyze curriculum, and utilizing cognitive
load theory to help reduce extraneous cognitive load while being
intentional about inclusion of germane cognitive load should help
instructional designers create learning environments in which students
are able to achieve learning goals. In some cases, intended goals may
have to be adjusted to account for the finite limitations of working
memory capacity: it is likely not feasible to expect students with no
programming experience to learn the syntax and grammar of a new
language \emph{and} higher level concepts associated with a paradigm
like OOP, \emph{and} be able to switch between abstraction layers in
the fluid manner necessary to solving complex problems with
programs. What this suggests is that closer coordination is needed
across courses in a curriculum so that goals of one become background
knowledge of subsequent courses, resulting in successful achievement
of program learning goals by the time of graduation.

>>>>

% sorting=none will order citations in order of appearance
% use sorting=nyt to sort by name, year, title
\usepackage[backend=biber,autolang=hyphen, bibencoding=inputenc, isbn=false, uniquename=false, sorting=none, sortcites=true, style=numeric, autocite=superscript, url=false]{biblatex} % biblatex setup
\addbibresource{computing.bib}
\appto{\bibfont}{\footnotesize} % for 10pt bibliography

@@

\item Even if goals were clearly stated, it is likely that many
  different approaches could be used successfully if chosen
  intentionally to align with the goals and measurements of success
  (assessment). In many of the studies, it is not immediately clear if
  the measures of success used are appropriate for the stated
  objectives. That is to say, the assessments used might not always be
  measuring all aspects of the desired learning objectives.
\end{enumerate}

@@

It should be strongly noted that we are not advocating for OOP as a
``best'' teaching paradigm. Rather, we want to point out that from
what has been reported, teaching OOP effectively can be
challenging. What many implementations seem to miss is that OOP
defines an entirely different paradigm for problem solving than is
used in procedural programming\autocite{kolling_problem_1999}. As
should not be surprising, there are some problems for which OOP
provides a good set of tools to solve in an efficient, elegant manner,
namely those with solutions that naturally map to real-world
objects\autocite{raymond_art_2003}. Different problems may be better
solved using a procedural, functional, or data-flow
paradigm\autocite{kay_history_1996,felleisen_functional_2009,crestani_experience_2010}. Ultimately,
our goal should be to help students analyze problem structures and
make informed decisions regarding what paradigm would be best for a
particular problem. What is not clear from the literature is whether
or not this consideration is made when designing problems for a
particular course. Indeed, in some cases there is evidence that OOP
concepts are added to a course ``on top of'' older content, without
necessarily changing the problem
sets\autocite{kolling_problem_1999,vilner_fundamental_2007}.

\subsection*{Cognitive load theory}
Cognitive load theory (CLT) is based on the premise that the capacity
of our working memory, where all conscious cognitive processing takes
place, is finite and relatively small: we are able to manage 2-3
interacting elements at a time. Examples of ``interacting elements''
in the current context would be recalling and using the syntax and
grammar of a particular programming language, reasoning about the
behavior of a single object in an OOP environment, or reasoning about
a solution to a particular high-level problem. CLT posits that in
light of this finite capacity of working memory, for learning to
occur, all aspects of a task must fit into this finite space. While
some contributors to cognitive load are necessary for the task at
hand, others are not\autocite{paas_cognitive_2003}. Cognitive load can
be divided into three different categories: \emph{intrinsic} cognitive
load is due to the complexity of the task itself and generally can not
be reduced without simplifying the task. \emph{Extraneous} cognitive
load is associated with work that is due to the instructional method,
but not part of the task at hand. For instance, if the educational
goal is to complete a programming task, but learners must search and
interpret reference material in order to do so, the search and
interpretation of this material would be considered extraneous
cognitive load. Finally, \emph{germane} cognitive load is also a
result of the instructional design, but unlike extraneous load,
germane load enhances the learning process. For instance, if a
learning goal is that students are able to navigate and use reference
material, than the previous example would be considered germane
cognitive load. What this example highlights is that whether
complexity due to the instruction design is extraneous or germane
depends on the specific learning goals. The role of the instructional
designer then is to be aware of intrinsic load, reduce extraneous load
and intentionally managing germane load.

\subsection*{Object oriented programming}
%% Reviewer: more explicit connections to cogsci
Alan Kay originally conceived of object oriented programming as
\emph{one} way to think about complex
systems\autocite{kay_history_1996}. The underlying metaphore of Kay's
conception for OOP was the biological cell: each object is designed to
accomplish a specific set of tasks such that the complexity of these
tasks is encapsulated within the object. Objects interact with one
another by sending and receiving messages, and systems of these
objects are assembled to solve complex
problems\autocite{kay_history_1996}.

@@

Ultimately, if our goal is to teach problem solving in general, we
would be doing a grave disservice to our students to leave them with
the impression that there is only one paradigm, whether it be OOP,
procedural, functional, or something not yet invented. As previously
mentioned, OOP and procedural paradigms are comprised of fundamentally
different concepts for problem solving, and so again is the functional
paradigm. Unsurprisingly, there is no paradigm that is the ``best''
fit for every problem. Leaving students with the belief that only one
paradigm exists makes it very difficult for them to learn how to
intentionally select and apply the ``best'' paradigm to solve a
particular problem. \citeauthor{kolling_problem_1999} argues that
performing a paradigm shift from procedural style to OOP is difficult,
and so concludes that if OOP is a desired outcome then it should be
taught \emph{first} since the reverse shift, from OOP to procedural,
is conceptually easier. What this recommendation neglects to account
for however is that both paradigms depend upon authoring programs in
some language, and this authoring process takes a certain amount of
cognitive capacity. To justify an objects-first approach we would have
to verify that the cognitive load associated with OOP concepts
themselves combined with the cognitive load of learning to author a
program in a particular environment, does not exceed the learner's
working memory capacity.

In a superficial way, CLT theory would suggest that OOP is a helpful
paradigm for managing cognitive complexity: objects encapsulate
complexity by hiding the details of implementation behind a simple
interface. For example, an object that represents a list of items may
have operations to add and remove items from the list. Conceptually we
can use this object with our common understanding of what it means to
``add'' an ``item'' to a ``list''. Critically, we do \emph{not} need
to be aware of how the adding and removal of the item is actually
implemented in code, or even how the storage is structured in memory
(e.g. an array or a linked-list). We refer to this use of CLT to
justify the reduction of complexity of OOP as superficial because it
focuses on only the conceptual paradigm itself without regard to
implementing it in some programming environment. In practice, we
usually ask our students to implement OOP concepts in a particular
programming language. That task requires not only understanding OOP as
a paradigm, but also working with the programming environment,
authoring a program, understanding the syntax and grammar of the
language, and most likely debugging it as well.

\subsection*{Cognitive challenges teaching OOP}
A behavior that characterizes expert programmers is the ability to
easily shift between different abstraction layers of a problem
space\autocite{armoni_teaching_2013}. This ability may be especially
crucial to becoming fluent in OOP as the concepts are somewhat further
removed from the actual syntax than in other paradigms. For example,
in the procedural paradigm the central concept is that of the
procedure, or subroutine. Subroutines are simply collections of
imperative commands grouped together under a label. Thus, if someone
can write and understand imperative code, the conceptual leap to
understanding the procedural paradigm is relatively small. In
contrast, to understand objects in OOP one must understand the concept
of state and methods operating on the state. Both state and method are
implemented with code (generally imperative), but to understand how an
object functions one has to imagine its initial state, and then how
each method call would change this state. Cognitive load theory would
suggest that understanding the dynamic interaction of state and method
calls is more challenging than grouping lines of imperative codes into
re-usable procedures. Thus to be fluent in OOP design, a programmer
\emph{must} be able to switch between thinking at the object layer and
the program layer. Teaching abstraction is best done by doing it by
example, while being explicit about the abstractions
used\autocite{armoni_teaching_2013}. However, we must keep in mind the
limited capacity of working memory: if the cognitive load associated
with writing code and thinking about state and methods fills up a
learner's working memory, the learner may have difficulty grasping
concepts of abstraction, even when the instructor explicitly points
them out. To reduce the cognitive load associated with writing code
and working with object state and methods, these tasks must be
practiced until somewhat automated: to a point at which they can be
done without conscious thought\autocite{ambrose_chapter_2010}.

\subsection*{Language choice}
Choice of language does matter: If the learning objectives of an
introductory programming course are to learn problem solving concepts
then a language with a simpler syntax is preferable to one with more
complex syntax features\autocite{koulouri_teaching_2014}. The
explanation for this can be found in CLT. Given the current example of
a goal to teach problem-solving techniques, there will be cognitive
load associated with thinking about different strategies one could
use. Learning the syntax rules of a programming language adds to the
cognitive load, but it does not aid in the process of learning
problem-solving techniques. In other words, learning the syntax of a
programming language is an extraneous cognitive load in this
context. Thus, it is in the educator's best interest to minimize
cognitive load associated with features that do not directly help with
the learning objectives of the course and select a teaching language
with easy-to-learn syntax.

@@

In theory, an iconic programming environment such as
Scratch\autocite{resnick_scratch_2009} or
Alice\autocite{cooper_using_2003} should aid in the learning of higher
level concepts, such as those associated with OOP, since graphical
languages eliminates the extraneous cognitive load of dealing with
syntax and grammar of a language. In fact, there has been success in
using iconic languages to teach first year programming
\autocite{calloni_iconic_1997}, and research indicates that teaching
graphical programming first does not inhibit a student from later
transferring learned high level concepts to textual programming
languages\autocite{hundhausen_can_2009}.

\citeauthor{snook_incorporation_2005} describe efforts to use the Alice
programming environment to teach introductory programming
concepts\autocite{snook_incorporation_2005}. Interestingly a later
report assessing the efficacy of the curriculum using Alice indicated
that while pre-post testing indicated learning gains with the
environment, focus group data indicated a dissatisfaction with Alice
as an introductory language, resulting in a switch to
LabVIEW\autocite{lohani_reformulating_2011}. A possible explanation
given for the dissatisfaction of Alice was that students did not
perceive it as being a ``real'' programming environment they might use
in industry, while LabVIEW was. Certainly, helping students connect
the utility of skills they learn in the classroom to those that will
be important in their later career is important. In some cases
switching to a tool that is known to be used in industry may be
useful, though this approach seems shortsighted: it is unlikely that
all graduates will end up in an industry that uses a particular tool
(e.g. LabVIEW), so if that is the motivation for a particular tool we
must still address the utility of this tool to students who may never
have reason to use it again. A different strategy than swathing tools
is to better articulate \emph{why} a particular tool was chosen for
the learning environment. Students might have less resistance to an
educational environment such as Alice if they know that it can help
them learn high level concepts that \emph{will} be directly applicable
to their future career \emph{and} that the use of Alice as an
introductory environment will not hinder them from later transferring
those concepts to other environments.

\subsection*{Problem solving}
Developing problem solving skills was widely regarded as a common objective
for introductory programming courses. Achieving this objective is more
likely if the course begins with a focus on problem solving
strategies, before any programming is
introduced\autocite{koulouri_teaching_2014}. Separating
problem-solving instruction in this way has a number of benefits
grounded in cognitive science research: first, it serves to activate
prior knowledge\autocite{ambrose_how_2010}. With problem-solving
strategies fresh in their memory, learners will more easily apply them
when they begin programming. Second, it helps manage cognitive
load. Where as in the previous example of language choice, we reduced
cognitive load by choosing a language with a simpler syntax, here we
separate two tasks that we know individually demand a high level of
cognitive load for novices: learning problem-solving strategies, and
learning programming techniques.

@@

\section*{Assessment}
%% Reviewer: short
Often times assessment becomes the weak point of a learning
system. Designing good assessment is
hard\autocite{lister_blooming_2000,traynor_synthesis_2005}, and often
good assessment can require significant time on the part of the
instructor. Because of this, assessment is often simplified, lowering
the effectiveness of the entire system.

If our course objective is to teach learners to solve problems with
programming, then an ideal assessment would be to give students a
problem to solve, ask them to write a program to solve it as well as
reflect upon the patterns they used in their solution. Asking for
reflection prompts students to engage in meta-cognition, becoming
aware of their own thinking process, which helps them become more
effective
learners\autocite{ambrose_how_2010,chi_selfexplanations_1989}. Of
course evaluating this type of assessment is much more time consuming
than a multiple-choice exam that can be automatically scored, or even
than evaluating that the finished program runs correctly, which can be
done automatically as well\autocite{cheang_automated_2003}.

\section*{Considerations and limitations}
In the above hypothetical example we had the luxury of complete
freedom in designing all aspects of an introductory programming
curriculum. We recognize this is almost always never the case in
practice; however, we argue that following a framework, such as
constructive alignment, as close as possible will help. Thinking about
the framework will also encourage critical engagement with the
constraints imposed by the institution. For example, there has been
much interest in assessment tools used for introductory programming
courses\autocite{lister_blooming_2000,lister_objectives_2001,sheard_exploring_2011},
and in particular, automated
assessment\autocite{cheang_automated_2003,traynor_synthesis_2005,ihantola_review_2010}. In
many cases the push towards automated assessment techniques is a
response to increasing class sizes that instructors typically have
little control over. In the event that a large class size dictates
that some form of automated assessment be used, following the
constructive alignment framework would constrain the types of course
objectives and instruction that would be feasible. For instance, if
the primary course objective was to teach conceptual understanding of
computer science, then a well-designed multiple-choice assessment
could be used
effectively\autocite{lister_blooming_2000,lister_objectives_2001}. Using
the framework of constructive alignment to critically evaluate the
alignment of assessment, objectives, and instruction, may help make
explicit some of the compromises that must be made for larger class
sizes. This positions course designers to be more intentional about
what compromises to learning are made.

@@

An area that has not been addressed in this paper but has strong
connections is that of attention in
cognition\autocite{simon_role_1986}. There are strong parallels
between our limitations to focus our conscious attention at one or two
tasks at a time, and the serial processing inherent in most computing
and language design. Of particular interest is the move in hardware to
increasingly parallel designs to keep pace with Moore's law. Our
cognitive constraints of attention may suggest an inherent difficulty
in thinking and reasoning about parallel processing that must be
addressed if we want to prepare students for an environment that
demands parallel-processing solutions to programming problems. These
demands may shift focus to languages such as Erlang that are designed
for high levels of concurrency, result in more interest in functional
paradigms over OOP, as well as demand we adopt different metaphors for
computing altogether\autocite{stein_challenging_1999}.

@@

Recognizing that introductory programming is a common component of an
engineering curriculum, and that there is a national push to introduce
it to an even larger population, it is important to think carefully
about how we design programming education to impart the cognitive
skills that will generally benefit learners even after the specific
technologies they learn are no longer relevant. We strongly advocate
for the use of a framework such as constructive alignment, described
in this paper, to inform the choices for course objectives,
instructional techniques, and assessments. Using such a framework well
help identify components of a course that may be contributing
unnecessarily to cognitive load, or reasons why student outcomes might
not match expectations assumed from course objectives.

Analyzing a proposed curriculum through the lens of cognitive load
theory can aid in instructional decisions that will increase the
chance of achieving learning goals. Educators must critically review
the goals of their introductory programming courses: if goals are to
understand high level concepts such as OOP, but not necessarily
implement these concepts in textual languages, it may be prudent to
consider iconic languages for these courses. If however, goals include
learning OOP concepts \emph{and} implementing these concepts in
textual language, educators must be aware of the increased cognitive
load associated with each of these tasks and provide scaffolding to
help students automate aspects of each task individually before
expecting them to combine both in successful programs.

Taken together, using an established framework, such as constructive
alignment, to design and analyze a curriculum, and utilizing cognitive
load theory to help reduce extraneous cognitive load while being
intentional about inclusion of germane cognitive load, should help
instructional designers create learning environments in which students
are able to achieve learning goals. In some cases, intended goals may
have to be adjusted to account for the finite limitations of working
memory capacity: it is likely not feasible in the span of a single
semester to expect students with no programming experience to learn
the syntax and grammar of a new language \emph{and} higher level
concepts associated with a paradigm like OOP, \emph{and} be able to
switch between abstraction layers in the fluid manner necessary to
solving complex problems with programs. What this suggests is that
closer coordination is needed across courses in a curriculum so that
goals of one become background knowledge of subsequent courses,
resulting in successful achievement of program learning goals by the
time of graduation.
